Sequential algorithms to split and merge terabyte-scale 3D images

### 2017-06-10 paper.tex
Changed model to count number of seeks instead of measuring total time

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

A disk is modeled by its read and write rates ($\alpha_r$ and
$\alpha_w$, in MB/s), and its access time ($\beta$, in seconds).
%also in
%MB/s).

@@

We adopt the following notations:
\begin{itemize}
\item $R$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $m$: amount of available RAM (in B).
\end{itemize}

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}$
\item Number of blocks required to fully cover a slice in the
  reconstructed image: $\sqrt[3]{n}^2=n^{\frac{2}{3}}$
  \item Number of full slices that can be reconstructed from $k$
    blocks: $\floor*{\frac{k}{n^{\frac{2}{3}}}}\sqrt[3]{\frac{R}{n}}$
\end{itemize}
\todo{Make a figure to illustrate the notations.}

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. This is due to the number of seeks performed by
Algorithm~\ref{algo:naive-blocks} to write a block, which equals the
total number of rows in all slices of a block, that is (since the
blocks are cubic): $\sqrt[3]{\frac{R}{n}}^2$. \note{Note the square
  which is omitted in equation 2 below and in model.py. Adding the
  square to the model just makes the runtime aberrant. This is because
  not all seeks are equal: only ``long'' seeks between block slices
  are significant; seeks between rows of a given block slice are much
  shorter. The way to model this is to replace $\beta$ by $\beta +
  \frac{s}{\gamma}$ where $\beta$ is in seconds, $\gamma$ is in B/s
  and $s$ is the seek ``distance'' in bytes. I bet $\gamma$ is very
  high on the SSD. } In practice, this could lead to a tremendous
slowdown as we will show later. For slices, the total merge time is
modeled as follows:
\begin{equation}
T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
\end{equation}
and for blocks:
\begin{equation}
  T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+\beta n\left(1+\sqrt[3]{\frac{R}{n}}\right)
\end{equation}

@@

The number of seeks is not reduced but the length of seeks is.

@@

@@

\begin{equation}
T_\mathrm{buff\_slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+\beta\left( n + \ceil*{\frac{bR}{m}}\right)
\end{equation}

@@

See Algorithm~\ref{algo:cluster-reads}. As many blocks as possible are
loaded in memory, concatenated in a buffer and written as sequentially
as possible from the buffer. Although the Algorithm looks very similar
to Algorithm~\ref{algo:buffered-slices}, it should be noted that in
Algorithm~\ref{algo:cluster-reads} the buffer should actually be a
more complex data structure capable of storing multiple disjoint
sequences of contiguous bytes without having to allocate memory to
store the ``holes'' between such sequences \todo{a figure showing the
  buffer and how it fillus up would help}.  \note{More on
  defragmentation might be needed here.}
\begin{algorithm}[h]
  \caption{Buffered merging from blocks using cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and append it to buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

\begin{equation}
  \begin{multlined}
    T_\mathrm{cluster\_reads} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w} +\\
    \beta
    \left[n+ 
      \ceil*{\frac{bR}{m}} \left[
        \sqrt[3]{\frac{R}{n}} \left(
        \ceil*{\frac{m\sqrt[3]{n}}{bR}} - \floor*{\frac{m\sqrt[3]{n}}{bR}}
        \right)+1
        \right]
      \right]
  \end{multlined}
\end{equation}

@@

     \begin{equation}
       T_\mathrm{multiple\_reads}= \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w} + \beta\ceil*{\frac{bR}{m}}\left(1+ \ceil*{\frac{m.n}{bR}}\right)\\
\end{equation}

>>>>

% packages
\usepackage{algorithm}
\usepackage{algorithmic}
\usepackage{amsfonts} % for R symbol (the set of real numbers)
\usepackage{color}
\usepackage[pdftex]{graphicx}
\usepackage{hyperref}
\usepackage{mathtools}
\DeclarePairedDelimiter{\ceil}{\lceil}{\rceil}
\DeclarePairedDelimiter{\floor}{\lfloor}{\rfloor}

@@

%also in
%MB/s).

@@

We adopt the following notations:
\begin{itemize}
\item $R=D^3$: number of voxels in the reconstructed image.
\item $b$: number of bytes per voxel in the reconstructed image and chunks (in B).
\item $n$: number of chunks (blocks or slices).
\item $\alpha_r$: data read rate from disk (in B/s).
\item $\alpha_w$: data write rate to disk (in B/s).
\item $\beta$: disk access time (in s).
\item $\sigma: \mathbb{R}^{+} \rightarrow \mathbb{R}^{+}$: seek time (in s), function of the seek
  distance (in B).
\item $m$: amount of available RAM (in B).
\end{itemize}

The following properties are straightforward applications of these
definitions:
\begin{itemize}
\item number of bytes per chunk: $\frac{Rb}{n}$
\item number of chunks that can fit in memory: $\frac{m.n}{Rb}$
\end{itemize}
For the sake of mathematical simplicity, the models below assume that
the reconstructed image and blocks are cubic. Under this assumption,
we have the following definitions and properties:
\begin{itemize}
\item Number of slices in a block: $\sqrt[3]{\frac{R}{n}}=d$. This is
  also the number of rows and columns in a slice and the number of
  voxel in a row or column.
\item Number of blocks required to fully cover a slice in the
  reconstructed image: $\sqrt[3]{n}^2=n^{\frac{2}{3}}$
  \item Number of full slices that can be reconstructed from $k$
    blocks: $\floor*{\frac{k}{n^{\frac{2}{3}}}}\sqrt[3]{\frac{R}{n}}$
\end{itemize}
\todo{Make a figure to illustrate the notations.}

\subsection{Disk model}

A disk is modeled by its read and write rates ($\alpha_r$ and
$\alpha_w$, in B/s), its access time $\beta$ (in s) and its seek time,
a function $\sigma$ of the seek distance (in B). More precisely,
$\sigma$ would typically be in the order of $\beta$ when an opened
file is seeked to a random position, but it would be much smaller when
a file is seeked to a nearby position.

The disk model encompasses file systems, kernel modules and disk
controllers required to access the disk from a program run in user
space. In modern systems, this often includes caching at various
levels \todo{\cite{x}}, which we will model by zeroing $\alpha_r$ or
$\alpha_w$ values depending on the application.

@@

Algorithms~\ref{algo:naive-slices} and~\ref{algo:naive-blocks}
actually have very different performance even though blocks and slices
have identical sizes. Since slices are stored contiguously in the
reconstructed image, the number of seeks in
Algorithm~\ref{algo:naive-slices} is only $2n$:
\begin{equation}
  N_\mathrm{slices} = 2n 
\end{equation}
However,
Algorithm~\ref{algo:naive-blocks} has to do extra seeks for each row
in each slice of each block. That is, the total number of seeks
performed by Algorithm~\ref{algo:naive-blocks} is:
\begin{equation}
N_\mathrm{blocks} = 2n+nd^2 = n\left(1+\sqrt[3]{\frac{R}{n}}^2\right)
\end{equation}
In
practice, this difference could lead to a tremendous slowdown as we will show
later.

%%  For slices, the total merge time is modeled as follows:
%% \begin{equation}
%% T_\mathrm{slices} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+2\beta n
%% \end{equation}
%% and for blocks:
%% \begin{equation}
%%   \begin{multlined}
%%     T_\mathrm{blocks} = \frac{bR}{\alpha_r}+\frac{bR}{\alpha_w}+ 2\beta n + \\
%%     n\left({\sigma(bR/2)} + d(d-1)\sigma\left((D-d)b\right)+d\sigma\left((D^2-d^2)b\right) \right)
%%     \end{multlined}
%% \end{equation}

@@

\todo{Not sure what we are measuring here. The number of seeks will be identical. The difference in seek time is supposed to be very minimal.}

@@

The algorithms presented up to now are in fact particular cases of
memory buffering where the amount of available memory equals the
maximum size of a chunk.

@@

\begin{equation}
N_\mathrm{buff\_slices} =  n + \ceil*{\frac{bR}{m}}
\end{equation}

@@

See Algorithm~\ref{algo:cluster-reads}. As many blocks as possible are
loaded in memory, concatenated in a buffer and written as sequentially
as possible from the buffer. Although the Algorithm looks very similar
to Algorithm~\ref{algo:buffered-slices}, it should be noted that in
Algorithm~\ref{algo:cluster-reads} the buffer should actually be a
more complex data structure capable of storing multiple disjoint
sequences of contiguous bytes without having to allocate memory to
store the ``holes'' between such sequences \todo{a figure showing the
  buffer and how it fillus up would help}.  \note{More on
  defragmentation might be needed here.}
\begin{algorithm}[h]
  \caption{Buffered merging from blocks using cluster reads}
  \label{algo:cluster-reads}
  \begin{algorithmic}
    \STATE list = sort blocks by increasing (x,y,z) values
    \STATE initialize buffer
    \STATE index = 0
    \WHILE{index\textless n}
      \WHILE{sizeof(buffer)\textless m}
        \STATE block = list[index]
        \STATE read block and insert it in buffer
        \STATE index += 1
        \ENDWHILE
       \STATE write buffer in reconstructed image
       \STATE clear buffer
    \ENDWHILE
  \end{algorithmic}
\end{algorithm}

\begin{equation}
  N_\mathrm{cluster\_reads}= n + \ceil*{\frac{bR}{m}}f(m,n,R)
\end{equation}
where:
\begin{equation}
  f(m,n,R)=
  \begin{cases}
    d^2 & $if there is an incomplete line$\\
    d   & $if there is an incomplete slice$\\
    1   & $otherwise$
\end{cases}
\end{equation}

@@

\begin{equation}
  N_\mathrm{multiple\_reads}= n \max\left(1,\ceil*{\frac{d}{\ceil*{\frac{m}{nd^2b}}}}\right)+ \ceil*{\frac{bR}{m}}
\end{equation}
